---
title: "Rugby World Cup 2015 Analysis"
author: "Michael Khan"
date: "4 January 2016"
output: html_document
---

----
Set working directory. 
Change this path as needed to run
```{r}
setwd("~/Dropbox/DataAnalytics/RugbyWC2015/code")
```


Necessary package imports
```{r}
library(reshape2)
library(cluster)
library(ggplot2)
```

###Importing Data
*data contains the original data provided.
*playerPosition contains data on players' physical attributes and positions
```{r}
data <- read.csv('../data/rwc2015.csv', header = TRUE)
playerPositions <- read.csv('../data/playerPositions.csv', header = TRUE)
```


###Quick exploration of the data
```{r}
dim(data)
names(data)
```
From this we can see that there are 4088 observations with 4 variables each: player_name, country, metric, value


Now lets look further into the data
```{r}
summary(data)
```
From this summary of the fields we can start to see how the data is organised in the file.
It looks like each player is represented by multiple rows of data, with the metric field showing which of the metrics is being show in this row.
```{r}
unique(data$metric)
```
As we can see from the above command there are 10 different metrics shown in this data.


###Data Munging
As I continue with my analysis I would prefer to transform this data so that each metric is a variable and each row is a player.
For this task I will use the [reshape2](https://cran.r-project.org/web/packages/reshape2/index.html) package imported above

```{r}
new_data <- dcast(data, player_name + country ~ metric)
new_data_NAs <- new_data
```

Here is an short extract of what the data look like now
```{r}
head(new_data)
```
As we can see from the above extract NA was inserted where the values did not exist for a given player and metric. It looks like the NAs signify when a player has 0 of a metric, although I am not 100% sure for this exercise I will replace NAs with 0
```{r}
new_data[is.na(new_data)] <- 0
```
The summary of the 10 different metrics' is now more true to reality
```{r}
summary(new_data)
```


I'm going to divide all the metrics by the number of games played by each player to remove bias of the number of games each player took part in.
```{r}
columnsToDivide <- c(3,4,5,6,7,9,10,11,12)
for(i in columnsToDivide) {
  new_data[ ,i] = new_data[ ,i] / new_data$matches
} 
```


Finding the leader in each category 
```{r}
max_carries <- new_data[new_data$carries == max(new_data$carries),]
max_cogl <- new_data[new_data$carries_over_gain_line == max(new_data$carries_over_gain_line),]
max_cb <- new_data[new_data$clean_breaks == max(new_data$clean_breaks),]
max_los <- new_data[new_data$lineout_steals == max(new_data$lineout_steals),]
max_low <- new_data[new_data$lineouts_won == max(new_data$lineouts_won),]
max_mm <- new_data[new_data$metres_made == max(new_data$metres_made),]
max_om <- new_data[new_data$offloads_made == max(new_data$offloads_made),]
max_tries <- new_data[new_data$tries == max(new_data$tries),]
max_tow <- new_data[new_data$turnovers_won == max(new_data$turnovers_won),]
```


###Clustering players

> I'm going to try to use K-means clustering to see if we can group players by positions by using just the 10 fields provided in the data. With K-means you have to specify the number of groups to use for clustering I'm going to start with 9, one for each position: prop, hooker, lock, back row, scrum half, fly half, center, wing and full back. 

The above quote shows my initial intention of using clustering algorithm, having some knowledge of Rugby I knew that k-means was unlikely to be able to create 9 separate clusters not due to a fault in the algorithm but because of the nature of the sport. Many players can interchange between multiple positions for example a player like Australia's Kurtley Beale can basically play any position in the back line. As well as players being interchangeable in modern day rugby the role of players in different positions can be very similar for example back-row forwards and centers can end up doing a lot of the same jobs and when looking at their stats it can be difficult to differentiate between position.
After a lot of experimentation with the number of clusters I decided that 3 clusters made the most sense, while this didn't split the players into each individual position it did create a pretty interesting division of the player. 
```{r}
set.seed(1)
kmeansIn <- data.matrix(new_data[, -c(1,2,8)])
kmeansRes <- kmeans(kmeansIn, centers = 3)
```

adding a new column to the data to hold the number of the cluster it was assigned to.
```{r}
new_data$cluster <- kmeansRes$cluster
```


####Merging player stats and position data.
To analyse the results of the clustering on player positions I had to find a data source that included player position.
I found a [dataset](http://www.rugbyhow.com/2015RWC-all-player-sizes.html#AName order) with physical measurements of players for the 2015 Rugby World Cup that also included player position. I merged the 2 data sets using player names 
```{r}
merged <- merge(x = new_data, y = playerPositions, by.x = 'player_name', by.y = 'Player.name')
merged_left <- merge(x = new_data, y = playerPositions, by.x = 'player_name', by.y = 'Player.name', all.x = TRUE)
merged_right <- merge(x = new_data, y = playerPositions, by.x = 'player_name', by.y = 'Player.name', all.y = TRUE)

missing_left <- merged_left[is.na(merged_left$Pos), ]
missing_right <- merged_right[is.na(merged_right$country), ]
```
only 535 of 620 players were matched. This issue seems to be caused mostly by small discrepancies in the names between the two data sources e.g. Apisalome Ratuniyarawa / Api Ratuniyarewa. For this exercise its ok to disregard the 87 mismatches, for other more important problems we could use a fuzzy string matching algorithm like the Levenshtein distance algorithm to match the strings that don't match exactly.

Lets look at a table of position vs cluster
```{r}
table(merged$cluster, merged$Pos)
```



###Cluster Plot
The hereunder silhouette plot provides an interesting way at visualizing the effectiveness of the clusters created by k-means in high-dimensional problem such as this one.  

Each horizontal bar represents the ratio of how well a player fits into his cluster as apposed to his next nearest cluster. The closer to 1 these bars are the better a player fits into his currency cluster.
Interesting observations from this graph include:
 *Cluster 3 is the largest and the best fit cluster with an average score ( s~i~ ) of 0.86
 *The small spike below 0 in cluster 1 signifies that a small number of players may have been more suitable in an other cluster. 
```{r}
graphColors = c('red', 'green', 'blue')
dE <- daisy(kmeansIn)
dE2 <- dE^2
silPlot <- silhouette(kmeansRes$cl, dE2)
plot(silPlot, col=graphColors, main='Silhouette plot of 3 player clusters')
```



###Exploratory Graphs

#####Comparing tries scored by cluster
```{r}
boxplot(merged$tries ~ merged$cluster, varwidth = TRUE, col=graphColors, xlab='Cluster', ylab='Tries scored per game', main='Tries scored per game cluster comparison')
ggplot(merged, aes(merged$tries, fill = as.factor(merged$cluster))) + geom_density(alpha = 0.2)
```


#####Comparing meters made by cluster
```{r}
boxplot(merged$metres_made ~ merged$cluster, varwidth = TRUE, col=graphColors, xlab='Cluster', ylab='Meters made per game', main='Meters made cluster comparison')
#ggplot(merged, aes(x=as.factor(merged$cluster), y=metres_made, fill=as.factor(merged$cluster))) + geom_boxplot() + scale_fill_discrete(name = 'Clusters')
ggplot(merged, aes(metres_made, fill = as.factor(merged$cluster))) + geom_density(alpha = 0.2) + scale_fill_discrete(name = 'Clusters') + ggtitle('Meters made cluster comparison')
```

#####Comparing turn overs by cluster
```{r}
boxplot(merged$turnovers_won ~ merged$cluster, varwidth = TRUE, col=graphColors, xlab='Cluster', ylab='Meters made per game', main='Meters made cluster comparison')
ggplot(merged, aes(turnovers_won, fill = as.factor(merged$cluster))) + geom_density(alpha = 0.2) + scale_fill_discrete(name = 'Clusters')
```

#####Comparing clean breaks by cluster
```{r}
boxplot(merged$clean_breaks ~ merged$cluster, varwidth = TRUE, col=graphColors, xlab='Cluster', ylab='Clean breaks made per game', main='Clean breaks made cluster comparison')
ggplot(merged, aes(clean_breaks, fill = as.factor(merged$cluster))) + geom_density(alpha = 0.2) + scale_fill_discrete(name = 'Clusters')
```

#####Comparing carries over gain line (cogl) by cluster
```{r}
boxplot(merged$carries_over_gain_line ~ merged$cluster, varwidth = TRUE, col=graphColors, xlab='Cluster', ylab='Carries over gain line per game', main='Carries over gain line cluster comparison')
ggplot(merged, aes(carries_over_gain_line, fill = as.factor(merged$cluster))) + geom_density(alpha = 0.2) + scale_fill_discrete(name = 'Clusters') + ggtitle('Carries over gain line cluster comparison')
```
